Critical assessment of Linear-Response Excitations in Molecular system\\ or \\ On the Localization of the Linear-Response excitations in Open Systems

### 2018-01-15 LRDraft.tex
update

\section{THEORY}

@@

\subsection{Localization properties of the response density}

We present an ulterior expression of the response density operator in the frequency space with the aim to establish its (spatial) localization properties.
The starting point of this analysis is the expression of the response density in terms of the commutator of \eqref{rhoPrimeOmegaDef1}. We observe that, for any 
operator $\op O$, we may write:  
\be
\commutator{\dmnot}{\op O} = \dmnot \op O \op Q_0 - \op Q_0 \op O \dmnot\;,
\ee
where $\op Q_0= \identity - \dmnot$ projects the operator in the subspace orthogonal to $\dmnot$. So, thanks to this relation we have that:
\be
\dm'(\omega) = \ii\sum_p\left(\ketbra{\psi_p}{\psi_p}\op A(\omega)\op Q_0 - \op Q_0\op A(\omega)\ketbra{\psi_p}{\psi_p}\right) \;.
\ee
This equation suggests us to define the \emph{fluctuation states}:
\be\lb{fluctuationStateDef1}
\ket{f_p(\omega)} \doteq -\ii\op Q_0\op A(\omega)\ket{\psi_p} \;, \qq
\bra{f_p(\omega)} \doteq \bra{\psi_p}\ii\op A(-\omega)\op Q_0 \;,
\ee
where the $\omega$ dependence of the bra is consistent with hermitian conjugation property of the Fourier transform of a hermitian operator. Fluctuation states are, by definition,
orthogonal to the occupied eigenstates of $\hnot$, so that:
\be
\brket{f_p(\omega)}{\psi_q} = \brket{\psi_q}{f_p(\omega)} = 0 \;, \forall p,q,\omega
\ee
The response density written in terms of the fluctuation states reads:
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;.
\ee
Applying the Liouville equation \eqref{LiouvillianRhopomegaDef3} to \eqref{rhoPrimeFluctuationStateDef1} allows us to derive a set of equation of motion for the
fluctuation states. The following equations are useful for this task:
\begin{align}
\Liouvnot\dm'(\omega) & = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)}\left(\eps_p-\hnot\right) + \left(\hnot-\eps_p\right)\ketbra{f_p(\omega)}{\psi_p}\right) \;, \nn \\
\coupl\dm'(\omega) &= \commutator{\op V'(\omega)}{\dmnot} = \op Q_0 \op V'(\omega) \dmnot - \dmnot \op V'(\omega) \op Q_0 = \nn \\
& = \sum_p\left( \op Q_0 \op V'(\omega) \ketbra{\psi_p}{\psi_p} - \ketbra{\psi_p}{\psi_p} \op V'(\omega) \op Q_0 \right) \;, \nn \\
\commutator{\op \Phi(\omega)}{\dmnot} & =  \sum_p\left( \op Q_0 \op \Phi(\omega) \ketbra{\psi_p}{\psi_p} - \ketbra{\psi_p}{\psi_p} \op \Phi(\omega) \op Q_0 \right) \;, \nn
\end{align} 
so that:
\begin{align}
\left[\bbomega - \left(\hnot-\eps_p\right)\right]\ket{f_p(\omega)} &= \op Q_0\left(\op V'(\omega) + \op \Phi(\omega)\right)\ket{\psi_p} \;, \nn \\
\bra{f_p(\omega)}\left[\bbomega + \left(\eps_p-\hnot\right)\right] &= \bra{\psi_p}\left(\op V'(-\omega) + \op \Phi(-\omega)\right)\op Q_0 \;.
\end{align}

>>>>

\section{General theoretical setup}

@@

\section{Critical Analysis of the linear response results}

The main aim of this section is to discuss the reliability in the assessment of the results provided by the linear response formalism. In particular we are interested in 
analyzing the \emph{linear response functional} defined as:
\be\lb{LinearResponseFunctDef1}
{\cal R}_{\op O}(\omega) = \trace{\dm'(\omega)\op O} \;,
\ee
that represents the perturbation-induced modification of the expectation value of an observable $\op O$. Here, for a sake of concreteness, we are limiting our considerations 
to observables that do not explicitly depend on time.

We start this analysis by deriving an ulterior expression of the response density operator in the frequency space. We consider equation \eqref{rhoPrimeOmegaDef1} and observe that, for any 
operator $\op O$, we may write:  
\be
\commutator{\dmnot}{\op O} = \dmnot \op O \op Q_0 - \op Q_0 \op O \dmnot\;,
\ee
where $\op Q_0= \identity - \dmnot$ is the projection operator in the subspace orthogonal to $\dmnot$. So, thanks to this relation we have that:
\be
\dm'(\omega) = \ii\sum_p\left(\ketbra{\psi_p}{\psi_p}\op A(\omega)\op Q_0 - \op Q_0\op A(\omega)\ketbra{\psi_p}{\psi_p}\right) \;.
\ee
This equation suggests us to define the \emph{fluctuation states}:
\be\lb{fluctuationStateDef1}
\ket{f_p(\omega)} \doteq -\ii\op Q_0\op A(\omega)\ket{\psi_p} \;, \qq
\bra{f_p(\omega)} \doteq \bra{\psi_p}\ii\op A(-\omega)\op Q_0 \;,
\ee
where the $\omega$ dependence of the bra is consistent with hermitian conjugation property of the Fourier transform of a hermitian operator. An inspection of equation \eqref{ketBraTimeEvol1} 
evidence that formulas \eqref{fluctuationStateDef1} are equivalent to the action of the $\op Q_0$ projector acting on the time evolution of the (occupied) eigenstates of $\hnot$, written 
in the $\omega$ domain. Fluctuation states are, by definition, orthogonal to the occupied eigenstates of $\hnot$, so that:
\be
\brket{f_p(\omega)}{\psi_q} = \brket{\psi_q}{f_p(\omega)} = 0 \;, \forall \, p,q,\omega \;.
\ee
The response density written in terms of the fluctuation states reads:
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;,
\ee
and consequently the linear response functional can be expressed as:
\be\lb{LinearResponseFunctDef2}
{\cal R}_{\op O}(\omega) = 
\sum_p\left(\bra{f_p(-\omega)}\op O\ket{\psi_p} + \bra{\psi_p}\op O\ket{f_p(\omega)}\right) \;.
\ee
The task of the linear response formalism to provided reliable results for the expectation value of $\op O$ can be reformulated in the problem of finding an accurate and unbiased
expression for the fluctuation states in the domain in which $\op O\ket{\psi_p}$ (and its hermitian conjugate) are not vanishing.

\subsection{Localization argument for the fluctuation states}

Starting from the Liouville equation \eqref{LiouvillianRhopomegaDef3} it is possible to derive a set of equation of motion for the fluctuation states. 
The following equations are useful for this task: 
\begin{align}
\Liouvnot\dm'(\omega) & = \sum_p\left(-\ketbra{\psi_p}{f_p(-\omega)}\left(\hnot-\eps_p\right) + \left(\hnot-\eps_p\right)\ketbra{f_p(\omega)}{\psi_p}\right) \;, \nn \\
\coupl\dm'(\omega) &= \commutator{\op V'(\omega)}{\dmnot} = \op Q_0 \op V'(\omega) \dmnot - \dmnot \op V'(\omega) \op Q_0 = \nn \\
& = \sum_p\left( \op Q_0 \op V'(\omega) \ketbra{\psi_p}{\psi_p} - \ketbra{\psi_p}{\psi_p} \op V'(\omega) \op Q_0 \right) \;, \nn \\
\commutator{\op \Phi(\omega)}{\dmnot} & =  \sum_p\left( \op Q_0 \op \Phi(\omega) \ketbra{\psi_p}{\psi_p} - \ketbra{\psi_p}{\psi_p} \op \Phi(\omega) \op Q_0 \right) \;. \nn
\end{align} 
Liouville equation written in this fashion implies the following equations for the fluctuation states, that can be extracted by multiplying from left and right for the 
occupied eigenstates of $\hnot$:
\begin{align}\lb{fluctuationStateEqMotion1}
\left[\bbomega - \left(\hnot-\eps_p\right)\right]\ket{f_p(\omega)} &= \op Q_0\left(\op V'(\omega) + \op \Phi(\omega)\right)\ket{\psi_p} \;, \nn \\
\bra{f_p(-\omega)}\left[-\bbomega - \left(\hnot-\eps_p\right)\right] &= \bra{\psi_p}\left(\op V'(\omega) + \op \Phi(\omega)\right)\op Q_0 \;.
\end{align}
These equations allows us to present a localization argument for the fluctuation states. We discuss this point for the right vectors, \emph{i.e.} for the first line of 
\eqref{fluctuationStateEqMotion1}, but the same argument holds also for the left ones.

We split the unperturbed Hamiltonian as the kinetic $\op T$ plus the potential $\op V$ term, so that the equation of motion for $\ket{f_p(\omega)}$ is recasted as:
\be\lb{fluctuationStateEqMotion}
\left(\bbomega+\eps_p -\op T\right)\ket{f_p(\omega)} = \op V\ket{f_p(\omega)} + \op Q_0\left(\op V'(\omega) + \op \Phi(\omega)\right)\ket{\psi_p} \;.
\ee

% We present an ulterior expression of the response density operator in the frequency space with the aim to establish its (spatial) localization properties.
% The starting point of this analysis is the expression of the response density in terms of the commutator of \eqref{rhoPrimeOmegaDef1}. We observe that, for any 
% operator $\op O$, we may write:  
% \be
% \commutator{\dmnot}{\op O} = \dmnot \op O \op Q_0 - \op Q_0 \op O \dmnot\;,
% \ee
% where $\op Q_0= \identity - \dmnot$ projects the operator in the subspace orthogonal to $\dmnot$. So, thanks to this relation we have that:
% \be
% \dm'(\omega) = \ii\sum_p\left(\ketbra{\psi_p}{\psi_p}\op A(\omega)\op Q_0 - \op Q_0\op A(\omega)\ketbra{\psi_p}{\psi_p}\right) \;.
% \ee
% This equation suggests us to define the \emph{fluctuation states}:
% \be%\lb{fluctuationStateDef1}
% \ket{f_p(\omega)} \doteq -\ii\op Q_0\op A(\omega)\ket{\psi_p} \;, \qq
% \bra{f_p(\omega)} \doteq \bra{\psi_p}\ii\op A(-\omega)\op Q_0 \;,
% \ee
% where the $\omega$ dependence of the bra is consistent with hermitian conjugation property of the Fourier transform of a hermitian operator. Fluctuation states are, by definition,
% orthogonal to the occupied eigenstates of $\hnot$, so that:
% \be
% \brket{f_p(\omega)}{\psi_q} = \brket{\psi_q}{f_p(\omega)} = 0 \;, \forall \, p,q,\omega
% \ee
% The response density written in terms of the fluctuation states reads:
% \be\lb{rhoPrimeFluctuationStateDef1}
% \dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;.
% \ee
